---
title: "NYC Airbnb Listing Price Prediction"
date: "2023-05-25"
author: "Kien Le, Tony Nguyen"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r}

```

<style type="text/css">
h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author {
  font-size: 18px;
  text-align: center;
}
h4.date { 
  font-size: 18px;
  text-align: center;
}

body {
  font-size: 14px;
}
</style>
```{r, echo=FALSE, message = FALSE, results = FALSE, warning=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(caTools)
```


## 1. Project Overview. 

Airbnb is an online marketplace and hospitality service platform that allows individuals to rent out their properties or spare rooms to guests seeking short-term accommodations. The Airbnb platform connects hosts and guests, providing a user-friendly interface where hosts can list their available spaces, set their own prices, and provide detailed descriptions and photos of their accommodations. Guests can then search for listings based on their desired location, travel dates, and specific preferences. The main purpose of this report is to analyze and predict the prices of Airbnb listings in New York City (NYC). This project is intended to provide valuable insights for both hosts and guests, enabling them to make informed decisions regarding pricing and booking accommodations in NYC.

Guiding questions:

1. What are the key factors that influence the pricing of Airbnb listings in NYC?

2. How accurate are the predictions of our model, and what are the potential limitations or uncertainties associated with the predictions?

Domain: Housing Price. The goal is to predict a continuous target variable (in this case, the price) based on a set of input variables (the other columns in the dataset). We will split the data set into a training set and a testing set, where the training set is used to train the model, and the testing set is used to evaluate its performance. Once we have built and evaluated the model, we can use it to predict the price of new Airbnb listings in the future based on their features. These are our initial thoughts and they might change later on

## 2. Data and Resources Used.

The data set we used is on Kaggle website and it is called the “New York City Airbnb Open Data”.The New York City Airbnb Open Data set contains information about Airbnb listings in New York City. Here is the list of the variables that we will mostly focus on:

1.`neighbourhood_group`: The borough or area of New York City where the listing is located.

Data Type: Categorical (string)

Example: "Brooklyn"

2.`room_type`: The type of room or accommodation.

Data Type: Categorical (string)

Example: "Private room"

3.`price`: The price per night to rent the listing.

Data Type: Numeric (integer)

Example: 149

4.`minimum_nights`: The minimum number of nights required for a stay.

Data Type: Numeric (integer)

Example: 1

5.`number_of_reviews`: The number of reviews received for the listing.

Data Type: Numeric (integer)

Example: 9

6.`availability_365`: The number of days the listing is available within a year.

Data Type: Numeric (integer)

Example: 365

7.`calculated_host_listings_count`: The total number of listings by the host.

Data Type: Numeric (integer)

Example: 6

## 3. Data Cleaning and Pre-processing


```{r include=FALSE}
data <- read.csv("airbnb_data.csv")
```


**Checking duplicated values:**
```{r}
sum(duplicated(data))
```


**Checking missing values:**
```{r}
sum(is.na(data))
```


**Dropping unnecessary columns which does not generate useful insights. The columns will be dropped are: `id`, `name`, `host_id`, `host_name`, `neighbor_hood`, `last_review`**
```{r}
data <- data [, -c(1,2,3,4,6,13)]
```


**Check how many levels in column `room_type` and `neighborhood_group`, and turn them into factors:**
```{r}
unique(data$room_type)
unique(data$neighbourhood_group)

data$neighbourhood_group <- factor(data$neighbourhood_group)
data$room_type <- factor(data$room_type)
```

**Fill out missing values and check to see if there is any missing values again:**
```{r}
data$reviews_per_month <- ifelse(is.na(data$reviews_per_month), 0, data$reviews_per_month)
sum(is.na(data))
```

**Price distribution**

The price range in NYC for airbnb is from 20 to more than 1000 per night, with the peak of about 100 dollars per night. I set the range of price to smaller than 1000 because the dataset is right-skewed so by doing that, it is easier to see the distribution.
```{r}
ggplot(data = data, aes(x = price)) + 
    geom_histogram(binwidth = 10) +
    labs(x = "Price ($)", y = "Number of Airbnb", title = "Price Distribution of Airbnb in NYC (Price < 1000)")+
    xlim(0, 1000)
```

In this dataset, most of the room type are "entire home/apt" and "private room". 
```{r}
#plot distribution of Room Types in NYC Airbnb Listings
ggplot(data = data, aes(x = room_type, fill = room_type)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label=..count..)) +
  labs(x = "Room Type", y = "Count", title = "Distribution of Room Types in NYC Airbnb Listings") +
  scale_fill_manual(values = c("#1f78b4", "#b2df8a", "#fb9a99")) 
```

There are a lot of outliers so I set the limit of y-axis to 1000 to have a better visualization of the dataset. We could see that the Shared room does not have much that is above 2000. However, the other have several that are approximately $10000
```{r}
#plot distribution of Room Types in NYC Airbnb Listings with price smaller than $1000
ggplot(data = data, aes(x=room_type, y=price, fill=room_type)) + 
  geom_boxplot(outlier.alpha=0.1) +
  labs(title="Distribution of Price by Room Type (Price < 1000)", x="Room Type", y="Price ($)") +
  scale_fill_manual(values=c("#1f78b4", "#b2df8a", "#fb9a99")) +
  ylim(0, 1000)

```


**Remove outliers**

Calculate IQR and quantile to remove outliers:
```{r}
quantile <- quantile(data$price, probs=c(.25, .75))
iqr <- IQR(data$price)
data_no_OL <- data %>% filter(price > (quantile[1] - 1.5*iqr) & price < (quantile[2] + 1.5*iqr))  
```

Plot the data set without the outliers
```{r}
ggplot(data = data_no_OL, aes(x = price)) + 
    geom_histogram(binwidth = 10) +
    labs(x = "Price ($)", y = "Number of Airbnb", title = "Price Distribution of Airbnb in NYC after removing outliers")
```



**Regression Model**

The linear regression model on data without outliers has an R-squared value of 0.4825, which means that the model explains approximately 48.25% of the variability in the target variable (price). 
Some of the coefficient can be interpreted as follow:
Airbnb listings with a private room have an expected price that is approximately $75.63 lower than listings with the reference room type (entire home/apt), holding all other variables constant.
Airbnb listings with a shared room have an expected price that is approximately $100.90 lower than listings with the reference room type (entire home/apt), holding all other variables constant.

```{r}
# Build a linear regression model using the no outliers data
model <- lm(price ~ neighbourhood_group + latitude + longitude + minimum_nights + room_type + number_of_reviews + availability_365 + calculated_host_listings_count, data = data_no_OL)
summary(model)
```


**Model with prediction**

First, we create a training set and test set. The training set has 80% of the data and the test set has 20%. THen we Create predictions fr the model. The printed result is the root mean squared error (RMSE) of the model and it is 49.84541, which means that the average difference between the predicted price and the actual price on the testing set is around $49. This indicates that the model is relatively accurate in predicting the prices of Airbnb listings based on the given features. 

```{r}
# Split the data into a training set and a testing set
set.seed(123)
split <- sample.split(data_no_OL, SplitRatio = 0.8)
train <- subset(data_no_OL, split == TRUE)
test <- subset(data_no_OL, split == FALSE)

# Use the model to predict prices on the testing set
predictions <- predict(model, newdata = test)

# Calculate the root mean squared error (RMSE) to evaluate the performance of the model
rmse <- sqrt(mean((test$price - predictions)^2))
print(rmse)

```
```{r}
# Fit the ANOVA model
model <- aov(reviews_per_month ~ room_type, data = data)

# Print the ANOVA table
summary(model)
```


